---
title: "Tutorial on Statistical Computing for Extremes with **R**"
author: "Léo Belzile"
subtitle: "Regression models for nonstationary extremes"
date: "June 30, 2023"
eval: true
cache: true
echo: true
format:
  revealjs:
    slide-number: true
    preview-links: auto
    theme: [simple, hecmontreal.scss]
    title-slide-attributes:
      data-background-color: "#002855"
bibliography: EVA2023-software.bib
---


```{r include=FALSE}

hecbleu <- c("#002855")
fcols <- c(gris = "#888b8d",
           bleu = "#0072ce",
           aqua = "#00aec7",
           vert = "#26d07c",
           rouge = "#ff585d",
           rose = "#eb6fbd",
           jaune = "#f3d03e")
pcols <- c(gris = "#d9d9d6",
           bleu = "#92c1e9",
           agua = "#88dbdf",
           vert = "#8fe2b0",
           rouge = "#ffb1bb",
           rose = "#eab8e4",
           jaune = "#f2f0a1")

knitr::opts_chunk$set(fig.retina = 3, collapse = TRUE)
options(digits = 3, width = 75)
```


## Nonstationarity

There are various forms of nonstationarity, including 

- trends, 
- time-varying variance,
- seasonality,  
- covariate effects.

No general theory

## Solution 1: preprocessing

- detrend data to approximate stationary.
   - for financial time series, fit ARMA-GARCH model
   - use Box-Cox transformation with regression model
- analyze residuals $R_i = Y_i - \widehat{f}(\mathbf{X}_i)$, where the postulated model is such that $\mathsf{E}(Y_i) = f(\mathbf{X}_i)$.

Benefit: more data to estimate trend

## Estimating probability of rare events

\begin{align*}
&\Pr(Y_i > y \mid \mathbf{X}_i) \\\quad &= \Pr\{R_i > y - \widehat{f}(\mathbf{X}_i)\mid \mathbf{X}_i\}
\\\quad & = \Pr\{R_i > y - \widehat{f}(\mathbf{X}_i)\mid \mathbf{X}_i, R_i > u\}\Pr(R_i > u) \\& \qquad + \Pr\{u  \leq R_i > y - \widehat{f}(\mathbf{X}_i)\mid \mathbf{X}_i\}\Pr(R_i \leq u)
\end{align*}
where the first term of the last line can be estimated using a generalized Pareto and the latter empirically (typically zero if $y$ is sufficiently large).

## Solution 2: hierarchical models

We can also incorporate parameters in the parameters of the extreme value distribution.

- Model $\mu=f_{\mu}(\mathbf{X}; \boldsymbol{\beta}_{\mu})$, $\sigma(\mathbf{X})=f_{\sigma}(\mathbf{X}; \boldsymbol{\beta}_{\sigma})$, etc.
- Typically keep shape constant.

For explanatories $\mathbf{X}$, fit regression models of the form
\begin{align*}
\mu(\mathbf{X}) = \beta_0 + \beta_1 \mathrm{X}_1 + \cdots \beta_p \mathrm{X}_p,
\end{align*}
and estimate parameters by maximum likelihood.

## Generalized additive models

Rather than linear effects or changepoints (if $\mathrm{X}_j$ is binary), we could consider smooths

\begin{align*}
f(\mathrm{X}_j) = \sum_{k=1}^K \beta_k b_k(\mathbf{X}_j)
\end{align*}
where $b_k$ are basis functions, e.g., thin-plate splines (`tp`), cubic regression splines (`cr`) and their cyclic counterparts.

## Parsimony is key

- There are few exceedances or extremes...
- The support restriction translates into up to $n$ inequality constraints: difficult to optimize.


Take home message: keep models simple!

## Penalized regression

Penalization to enforce sparsity or regularize estimation.

- For generalized additive models (GAM), penalize square of second derivative to control the wiggliness of the function.
- Smooth matrix $\mathbf{S}$ with entries $S_{ij} = \int b_i''(x) b_j''(x)\mathrm{d} x$, so the penalty is
$\lambda \boldsymbol{\beta}^\top \mathbf{S}\boldsymbol{\beta}$ for $\lambda > 0$.
- These correspond to improper (i.e., rank-deficient) Gaussian priors.

## Tuning parameter

Maximization over
\begin{align*}
\max_{\lambda  \in \mathbb{R}_K^{+}} \max_{\boldsymbol{\beta}} \ell_{\boldsymbol{\lambda}}(\boldsymbol{\beta}) - \frac{1}{2} \boldsymbol{\beta}^\top\mathbf{S}_{\boldsymbol{\lambda}}\boldsymbol{\beta}
)
\end{align*}
where $\mathbf{S}_{\boldsymbol{\lambda}} = \sum_{k=1}^K \lambda_k \mathbf{S}_k$
-  Optimal value of $\boldsymbol{\lambda}$ from maximizing the marginal likelihood, obtained by integrating out $\boldsymbol{\beta}$ using Laplace's method [@Wood.Pya.Safken:2016].


## A word of caution

- With time-varying covariates, we need forecasts to compute risk measures in future.
- Model often fitted only at specific covariate values (careful at extrapolation beyond range).
- Rather than linear trend in time, use covariates that incorporate said trends but are more natural (e.g., climate model output).


## Return levels under nonstationarity

In nonstationary models, risk measures of interest are defined conditionally on the value of covariates

For example, the $1-p$ conditional return level is [@Eastoe.Tawn:2009]
\begin{align*}
\Pr(Y_t  > y \mid \mathbf{X}_t =\boldsymbol{x}_t) = p
\end{align*}

## Unconditional return levels

These can be obtained by averaging out over the distribution of covariates that are employed in the model.

\begin{align*}
\int_{\mathcal{X}} \Pr(Y_t  > y \mid \mathbf{X}_t =\boldsymbol{x}_t) \mathrm{d} P(\boldsymbol{x}_t),
\end{align*}

Return levels may be meaningless quantities.

May require information about future distribution of covariates if these are time-varying.

## Threshold stability

The generalized Pareto model with varying scale and shape is not threshold-stable unless, for any $v>u$, 
\begin{align*}
\sigma_v(\boldsymbol{x}_t) = \sigma_u(\boldsymbol{x}_t) + (v-u) \xi(\boldsymbol{x}_t)
\end{align*}

Restrictive! For constant $\xi$, need $\sigma$ linear or constant (log scale) [@Eastoe.Tawn:2009].

Using the inhomogeneous Poisson point process representation avoids these problems.

For nonstationary threshold models $u(\mathbf{X})$, see Section 3.2.2 of @Northrop.Jonathan:2011.

## Software

Function `evgam` from the eponymous package [@Youngman:2022] builds on the `mgcv` package [@Wood:2017:mgcv].

The setup is `evgam(formula, data, family, ...)`, where 

- `family` is the character string for the extreme value distribution (`gev`, `gpd`, `rlarg` and `ald` for asymmetric Laplace, used in quantile regression),
- `formula` is a list of formula for parameters (in the order location, scale, shape).

## Formulas and smooth

Scale input first

Use `s` for smooths with arguments

- `k` controls the number of breakpoints
- `bs` controls the basis function
- `fx`: control whether fixed degrees of freedom for regression spline or else penalized regression spline (default, `FALSE`)

and `te` for tensor products of smooths (interactions)

## Data setup
```{r}
#| label: evgam-setup
library(evgam)
library(lubridate)
data(frwind, package = "mev")
lyon <- with(frwind,
             xts::xts(x = S2, order.by = date))
ymax <- xts::apply.yearly(lyon, max)
ymax <- ymax[-length(ymax)] # Remove 2023
```


## Example with `evgam`


```{r}
opt_gev <- evgam::evgam(
  data = data.frame(
    year = year(ymax),
    ymax = ymax),
  formula = list(ymax ~ s(year, bs = "cr"),
                 ~ 1, 
                 ~ 1),
  family = "gev")
```

# `evgam` output

```{r}
## Summary with coefficients
summary(opt_gev)
```

## Fitted splines 

```{r}
#| label: fig-splines
#| fig-cap: "Fitted spline for location parameter of generalized extreme value distribution as a function of day of year (scaled to unit interval)."
#| out-width: '60%'
plot(opt_gev)
```
Plot and generalized analysis of deviance suggests constant location.

## Estimated quantiles

For given $p$, solve nonlinear equation $$\prod_{i=1}^n F^N(z; \boldsymbol{\theta}_i, \mathbf{X}_i) = p$$ using `qev` to give unconditional quantile.

```{r}
#| eval: false
#| echo: false
# Careful! - default is type='link' (log scale)
sim_post <- simulate(object = opt_gev, 
                     nsim = 1000L,
                     type = "response")
# Simulated parameter for each combo
# averaged across observed values of 
# covariates in original data frame.
retlev <- with(sim_post, 
               evgam::qev(p = 1-1/50,
           loc = location,
           scale = scale,
           shape = shape,
           family = "gev"))
mean(retlev)
```


# Return levels (direct method)

```{r}
# More straightforwardly
mean(simulate(opt_gev, 
              nsim = 1000L, 
              type = "quantile", 
              probs = 1-1/50))
```

These return levels are meaningless, since time increases (but would lead to extrapolate beyond range of observed time)...

## Quantile regression

We can work the asymmetric Laplace distribution to build a working likelihood for quantile regression.

The latter has density function
\begin{align*}
\frac{\tau(1-\tau)}{\sigma} \exp\left\{-\rho_\tau \left(\frac{y-u}{\sigma}\right)\right\},
\end{align*}
where $\rho_\tau(y) = y(\tau-\mathrm{I}_{y <0})$ is the check function.

## Time-varying threshold with quantile regression

```{r}
qreg <- evgam::evgam(
  # Cubic cyclic splines
  formula = list(wind ~ s(tday, bs = "cc"), 
                 ~ s(tday, bs = "cc")), 
  data = data.frame(wind = as.numeric(lyon),
                    tday = yday(lyon)),
  family = "ald", 
  ald.args = list(tau = 0.95))
```

## Temporal trend

```{r}
plot(qreg)
```
# Extract exceedances

```{r}
u <- fitted(qreg)[,'location']
df <- data.frame(exc = as.numeric(lyon - u),
                  yday = yday(lyon),
                 u = u)
df <- df[lyon >u, ]
```



```{r}
opt_gpd <- evgam::evgam(
  formula = list(exc ~ 1, ~1),
  data = df,
  family = "gpd")
post_sim <- simulate(opt_gpd, 
                     type = "response")
```

# Return levels

```{r}
# Number of exceedances per year
npy <- diff(range(date(lyon))) / nrow(df)
retlev_gpd <- qev(
  p = 1-1/50,
  m = as.numeric(npy),
  loc = df$u,
  scale = post_sim$scale,
  shape = post_sim$shape,
  family = "gpd")
mean(retlev_gpd)
```

<!-- allow breaks over multiple slides -->

## References
