---
title: "Tutorial on Statistical Computing for Extremes with **R**"
author: "Léo Belzile"
subtitle: "Bayesian inference"
date: "June 30, 2023"
eval: true
cache: true
echo: true
format:
  revealjs:
    slide-number: true
    preview-links: auto
    theme: [simple, hecmontreal.scss]
    title-slide-attributes:
      data-background-color: "#002855"
bibliography: EVA2023-software.bib
---


```{r include=FALSE}

hecbleu <- c("#002855")
fcols <- c(gris = "#888b8d",
           bleu = "#0072ce",
           aqua = "#00aec7",
           vert = "#26d07c",
           rouge = "#ff585d",
           rose = "#eb6fbd",
           jaune = "#f3d03e")
pcols <- c(gris = "#d9d9d6",
           bleu = "#92c1e9",
           agua = "#88dbdf",
           vert = "#8fe2b0",
           rouge = "#ffb1bb",
           rose = "#eab8e4",
           jaune = "#f2f0a1")

knitr::opts_chunk$set(fig.retina = 3, collapse = TRUE)
options(digits = 3, width = 75)
```


## Bayesian inference

- Combine likelihood with a prior: posterior.

$$
 p(\boldsymbol{\theta} \mid \boldsymbol{Y}) = \frac{p(\boldsymbol{Y} \mid \boldsymbol{\theta})p(\boldsymbol{\theta})}{ \int p(\boldsymbol{Y} \mid \boldsymbol{\theta})p(\boldsymbol{\theta})\mathrm{d}\boldsymbol{\theta}},
$$ {#eq-bayes}

- Posterior is proper (if likelihood times prior is integrable).
- Parameters asymptotically normal in large samples under general conditions.
- Need to evaluate integral in denominator (normalizing constant that does not depend on $\boldsymbol{\theta}$).


## Change of paradigm

In the Bayesian world,

- inference is conditional on observations, 
- parameters are treated as random variables
   - (reflecting uncertainty about their true value).

## Inference

- In all scenarios, we will have samples from the posterior distribution for parameters $\boldsymbol{\theta}$: any summary can be computed from these via Monte Carlo methods.

- e.g., we can report the posterior mean or median
- credible intervals (probability that parameter falls in) are simply quantiles of functionals

## Prediction 

We consider new realizations from **posterior predictive**
\begin{align*}
p(Y_{\text{new}} \mid \boldsymbol{Y})
\int_{\Theta} p(Y_{\text{new}}  \mid \boldsymbol{\theta}) p(\boldsymbol{\theta} \mid  \boldsymbol{Y}) \mathrm{d} \boldsymbol{\theta}
\end{align*}

- For every posterior draw $\boldsymbol{\theta}_b$ $(b=1, \ldots, B)$, sample new realization from generative model.
- Marginalization means discarding $\boldsymbol{\theta}_b$'s

## Why bother?

- Priors can act as penalty
   - regularizing parameter estimates in small samples.
- Natural extensions via hierarchical modelling (e.g., random effects).
- Expert knowledge can be encoded through the priors.


## Intractable models

Trade optimization for integration problem!


The integral is intractable, so we need to resort to alternative schemes:

- Numerical integration via Laplace approximations (INLA)
- Monte Carlo methods to simulate (approximate) draws from the posterior

## Monte Carlo methods

- Most packages implement offer Markov chain Monte Carlo algorithms
   - gives approximate correlated draws
   - requires tuning
   - need to check convergence
   
   
For models without covariates, @revdbayes offers *exact* draws from posterior via the ratio-of-uniform algorithm.
   
## Bayesian dispute

One objection to Bayesian inference is arbitrary selection of priors

- Influence of the prior vanishes as sample size grows unless it imposes support constraints!
- Likelihood are also often used for convenience
- Same for lasso and ridge ($L_1$ and $L_2$) penalties in regression.

## Prior selection

- Typically, parameters are assumed independent apriori (so prior factorizes), even if posterior doesn't
- Choosing proper priors (density functions) ensure proper posteriors.
- Care is needed, especially for shape parameter

## Priors for extreme value models (1/2)

Popular choices include

- improper uniform priors
- vague or diffuse Gaussian priors, e.g., $\xi \sim \mathsf{No}(0, 10)$
- Jeffrey's prior (invariant to reparametrization.
   - for generalized Pareto $$\sigma^{-1}(1+\xi)^{-1}(1+2\xi)^{-1/2}\mathrm{I}_{\xi > -1/2}$$
   - invalid for the GEV regardless of sample size[@Northrop.Attalides:2016]!

## Priors for extreme value models (2/2)

Better choices penalize the shape, including

- @Martins.Stedinger:2001 prior $\xi + 0.5 \sim \mathsf{Beta}(6, 9)$
- Maximal data information (MDI): $\sigma^{-1}\exp\{-a(1+\xi)\}\mathrm{I}_{\xi \geq -1}$ @Northrop.Attalides:2016
- Prior on quantiles [@Coles.Tawn:1996], back-transformed

## Impact of priors

```{r}
#| cache: true
#| eval: true
#| echo: false
#| label: fig-posteriorxi
#| fig-cap: "Marginal posterior of GEV shape parameter for different prior distributions."
#| fig-align: 'center'
#| message: false
library(revdbayes)
library(ggplot2)
data("frwind", package = "mev")
lyon <- with(frwind,
             xts::xts(x = S2, order.by = date))
# Create series of yearly maximum
ymax <- as.numeric(xts::apply.yearly(lyon, max))
# Fit a model with a trivariate normal prior for mu, log(sigma), xi
prior1 <- set_prior(prior = "mdi", model = "gev")
prior2 <- set_prior(prior = "beta", model = "gev")
prior3 <- set_prior(prior = "norm", 
                model = "gev", 
                mean = c(mean(ymax), log(sd(ymax)), 0), 
                cov = diag(c(1000, 1000, 1)))

post_1 <- revdbayes::rpost_rcpp(
  n = 1e4L, 
  model = "gev",
  data = ymax,
  prior = prior1,
  nrep = 100)
post_samp1 <- post_1$sim_vals
post_samp2 <- revdbayes::rpost_rcpp(
  n = 1e4L, 
  model = "gev",
  data = ymax,
  prior = prior2)$sim_vals
post_samp3 <- revdbayes::rpost_rcpp(
  n = 1e4L, 
  model = "gev",
  data = ymax,
  prior = prior3)$sim_vals
# Compute marginal posterior for shape
ggplot(data = data.frame(
  shape = c(post_samp1[,'xi'],
            post_samp2[,'xi'],
            post_samp3[,'xi']),
  prior = rep(c("mdi", "beta", "normal"), 
              each = 1e4L)),
  mapping = aes(x = shape,
                col = prior, 
                group = prior)) +
  geom_density() +
  scale_color_viridis_d() + 
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Ratio-of-uniform sampling 

The `revdbayes` [@revdbayes] provides the state-of-the-art for stationary models (faster than MCMC, exact sampling).

```{r}
#| eval: true
#| echo: false
library(ggplot2)
data("frwind", package = "mev")
lyon <- with(frwind,
             xts::xts(x = S2, order.by = date))
ymax <- as.numeric(xts::apply.yearly(lyon, max))
```

```{r}
#| eval: true
#| echo: true
library(revdbayes)
post <- revdbayes::rpost_rcpp(
  n = 1e4L, # number of posterior samples
  model = "gev", # extreme value distribution
  data = ymax, # vector of yearly maximum
  prior = set_prior(prior = "mdi", model = "gev"),
  nrep = 100) # number of post. predictive samples
```

## Point estimators and credible intervals

For each combination of posterior draw, compute the functional of interest (e.g., the median of the 50-year maximum) and/or quantiles.

```{r}
post_gev_Nmed <- apply(post$sim_vals, 1, function(x){
  revdbayes::qgev(p = 0.5, loc = x[1], scale = x[2], 
                  shape = x[3], m = 50)
})
# Posterior mean
mean(post_gev_Nmed)
# To get a 95% credible interval, simply compute quantiles
quantile(post_gev_Nmed, c(0.025, 0.975))
```

## Posterior predictive samples

Here, using max-stability to get a new observation (maximum of 50 yearly max) for each draw $\boldsymbol{\theta}$.

```{r}
postpred <- revdbayes::rgev(
  n = nrow(post$sim_vals), 
  loc = post$sim_vals[,1],
  scale = post$sim_vals[,2],
  shape = post$sim_vals[,3],
  m = 50)
```

## Posterior predictive


```{r}
#| label: fig-postpred
#| echo: false
#| fig-cap: "Density of posterior predictive (black) and posterior median (grey) of 50 year maximum."
ggplot(data = data.frame(
  postpred = postpred,
  postmed = post_gev_Nmed)) +
  geom_density(mapping = aes(x = postmed), color = "grey") +
  geom_density(mapping = aes(x = postpred), color = "black", linewidth = 2) +
  labs(x = "average wind speed (in km/h)") + 
  theme_classic() +
  theme(legend.position = "none") 
```

## Diagnostics for Bayesian workflow

Can model capture summaries of data?

```{r}
pp_check(post, stat = median)
```

See `bayesplot` and `loo` packages for more information and options about the Bayesian workflow [@Gabry:2019].

## Regression model

```{r}
library(texmex)
post_reg <- texmex::evm(
  y = ymax, 
  data = data.frame(
    syear = scale(1976:2023), 
    ymax = ymax),
  family = texmex::gev,
  method = "simulate",
  burn = 1000L,
  chains = 4L,
  iter  = 1.1e4,
  proposal.dist = "gaussian",
  mu = ~ syear,
  verbose = FALSE)
```

## Posterior estimates

```{r}
summary(post_reg)
```
```{r}
#| eval: false
## Default 'plot' method for texmex objects:
# density, traceplots and correlograms plots
library(gridExtra)
ggplot(post_reg)
```

## Monitoring Markov chain Monte Carlo

- run multiple Markov chains from different starting values
- discard initial samples (burn-in)  
- check convergence:
   - are all chains converging to the same region 
   - are chains stationary?

## Effective sample size

With multiple chains, it is easier to use output of `coda` package.

For summaries to be reliable (e.g., quantiles of posterior), the approximate number of independent samples from the total simulated should be large enough.
```{r}
chains <- coda::as.mcmc.list(
  lapply(post_reg$chains, coda::as.mcmc))
# effective sample size is sufficient here
coda::effectiveSize(chains)
```

Can compare algorithms efficiency via effective sample size per second.

## Traceplots

```{r}
## Default 'plot' method for texmex objects:
# density, traceplots and correlograms plots
plot(chains, density = FALSE)
```

Should look like fat hairy catterpilars.


## Algorithm efficiency

While some MCMC algorithms are more costly, they yield samples that are less autocorrelated (so contain more information altogether).

- Is autocorrelation reasonable?
- Related to acceptance rate (Goldilock principle)
- Ideally, algorithms sample parameters on unconstrained space 
- Good proposals adapt to the local geometry and satisfy support constraints

## Correlograms
```{r}
coda::acfplot(chains)
```
## Summaries

We can get directly summaries (e.g., posterior sample mean)

- standard errors for functionals is more complicated due to the autocorrelation

Due to autocorrelation, dedicated methods are employed to get standard errors.

```{r}
summary(chains)
```

## Loss function

Custom loss function for the return levels $q$ from EVA data challenge
\begin{align*}
L(q, \widehat{q}(\theta)) = 
\begin{cases}
0.9(0.99q - \widehat{q}), & 0.99q > \widehat{q} \\
0, & |q - \widehat{q}| \leq 0.01 q\\
0.1(\widehat{q} - 1.01q), & 1.01q < \widehat{q}.
\end{cases}
\end{align*}

## Bayesian evaluation of loss function

For given value of the return level $q_0$:
\begin{align*}
 r(q_0) = \int_{\boldsymbol{\Theta}}L(q(\boldsymbol{\theta}), q_0) p (\boldsymbol{\theta}) \mathrm{d} \boldsymbol{\theta}
\end{align*}
and we seek to minimize the risk, $\mathrm{min}_{q_0 \in \mathbb{R}_{+}} r(q_0)$

## Loss function

```{r}
#| label: fig-lossfn
#| out-width: '80%'
#| eval: true
#| echo: false
#| fig-align: 'center'
#| fig-cap: "Loss function for wind speed data, based on fitting a generalized extreme value distribution to annual maxima."
gev_retlev <- function(par, N, p = 0.368){
  # Map parameters via GEV max-stability
  mu <- par[1] + par[2]*(N^par[3]-1)/par[3]
  sigma <- par[2]*N^par[3]; 
  xi <- par[3]
  # quantile of N-block maximum
  mev::qgev(p = p, loc = mu, scale = sigma, shape = xi)
}

# Loss function
loss <- function(qhat, q){
    mean(ifelse(0.99*q > qhat,
           0.99*(0.99*q-qhat),
           ifelse(1.01*q < qhat,
                  0.1*(qhat-1.01*q),
                  0)))
}
# Compute the posterior of the return levels
retlev_post <- apply(post_samp1, 1, gev_retlev, N = 50)
# Create a grid of values over which to estimate the risk
retlev_psi <- seq(
  from = quantile(retlev_post, 0.2),
  to = quantile(retlev_post, 0.99), 
  length.out = 101)
# Create a container to store results
risk <- numeric(length = length(retlev_psi))
for(i in seq_along(risk)){
  # Compute integral (Monte Carlo average over draws)
 risk[i] <- loss(q = retlev_post, qhat = retlev_psi[i])
}
# Plot loss function
ggplot(data = data.frame(
  loss = risk, 
  retlev = retlev_psi), 
  mapping = aes(x = retlev, y = loss)) +
  geom_line() +
  geom_vline(xintercept = mean(retlev_post)) +
  labs(x = "return level") +
  theme_minimal()
```

## References
